---
title: "STAT 410 Group Research Project Part 1: Proposal"
output: html_document
author: Sam Minard & Efrain Ramirez
---

First, I am going to explore a data set from Pew Research Center (American Trends Panel Wave 124). It is a .sav file, which requires the "haven" package to read it in. <https://www.pewresearch.org/dataset/american-trends-panel-wave-124/>

```{r setup}
#install.packages("haven")
library(haven)
library(readxl) 
library(tidyverse)
library(ggplot2)
```

```{r load-data}
#| echo: false
#dataset is called American Trends Panel Wave 124
#load the data
amertrends <- read_sav("data/data-raw/W124_Mar23/ATP W124.sav")

#load the codebook for the data to understand what each variable represents
amertrends_codebook <- read_excel("data/data-raw/W124_Mar23/ATP W124 Codebook.xlsx")
#view(amertrends_codebook)

```

### **Potential research questions we can investigate:**

1.  What demographic or political characteristics are associated with support for a strong leader who rules without checks and balances (i.e. autocracy)? (dependent variable: POLSYS_AUTOC_W124)

2.  How does perceived economic hardship (ECON_SIT_W124) relate to preferences for authoritarian leadership? (dependent: POLSYS_AUTOC_W124, POLSYS_JUNTA_W124)

3.  Which demographic or political factors predict satisfaction with the way democracy is working in the US? (SATISFIED_DEMOCRACY_W124)
